---
title: "Equations for object detection and moving speed by walk and wheelchair"
subtitle: "A sunday study in Chiang Mai, Thailand"
author: "Yuzuru Utsunomiya"
date: "`r Sys.Date()`"
format:
  pdf:
    # include-in-header: 
    #   text: '\usepackage{lineno}\linenumbers'
    pdf-engine: xelatex
    number-sections: false
    toc: false
    documentclass: article
    mainfont: "TeX Gyre Termes"
    linestretch: 2
    header-includes:
      - \usepackage{xeCJK}
      - \setCJKmainfont{Noto Serif CJK JP}
      - \usepackage{lineno}\linenumbers
keep-tex: true
crossref:
  eq-prefix: "eq."
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---


```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, results='hide'}
knitr::read_chunk("moving_speed_temples_cnx.r")
```

```{r read.library, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, results='hide'}
```

```{r table1, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, results='hide'}
```


# Introduction

## Introduction


## Prior study


## Materials and methods

### Target area


```{r tab1}
#| label: tbl-tableone
#| tbl-cap: "Summary Table stratified by mode and occasion"

as_kable_extra(
  object_time_tableone,
  format = "latex"
  ) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::kable_styling(font_size = 8)


```

### Model to estimate moving speed 

In this study, we adopted a state-space modeling framework to account for temporal fluctuations in walking pedestrians' and wheelchair users' movement speed. State space model [@commandeurIntroductionStateSpace2007].  Specifically, we formulated a dynamic regression model with an AR(1) latent process and interaction effects. The model consisted of two components: a state equation that describes the evolution of an unobserved, time-varying latent trend (@eq-state), and an observation equation that links this latent trend to the observed movement speed along with covariates and interaction effects (@eq-obs). The latent trend was modeled as a first-order autoregressive process (AR(1)), capturing unobserved, smooth temporal shifts in mobility behavior, while the observation equation incorporated the effects of mode, occasion, traffic components, and spatial variables, including an interaction term between mode and the number of cars (@eq-mu). This modeling strategy enabled us to separate short-term temporal variability (via the latent trend) from structural covariate effects, providing a more nuanced interpretation of mobility dynamics in the target area.

The AR(1) + interaction model is specified as follows:

$$
x_t = \phi_1 x_{t-1} + \epsilon_t, \quad \epsilon_t \sim \mathcal{N}(0, \sigma_x^2)
$$ {#eq-state}

where $x_{t}$ denotes latent trend of the state; $\phi$ denotes coefficient of burabura, indicating burabura; $\epsilon_{t}$ denotes error term of the ratent trend; and $\sigma$ denotes standard deviation of the state space.
<!-- Meaning? -->

We assume that the observation process follows a Student's-t distribution $Student's\ t(\nu,\ \mu,\ \sigma)$ given,

$$
y_t \sim \text{Student}_t(\nu = 4, \mu_t, \sigma_{y})
$$ {#eq-obs}

where $y_{t}$ denotes observed moving speed (km/h); $\mu_{t}$ denotes mean of the observed speed; and $\sigma_{y}$ denotes standard deviation of the observed space. The $\mu_{t}$ is decomposed in the following given,


$$
\begin{aligned}
\mu_t =\ & \alpha + x_t \\
& + \beta_{\text{car},\, j[t]} \cdot \log(\max(N_{\text{car},t}, 10^{-3})) \\
& + \beta_{\text{person}} \cdot \log(\max(N_{\text{person},t}, 10^{-3})) \\
& + \beta_{\text{motorcycle}} \cdot \log(\max(N_{\text{motorcycle},t}, 10^{-3})) \\
& + \beta_{\text{lon}} \cdot \text{lon}_t + \beta_{\text{lat}} \cdot \text{lat}_t \\
& + \alpha^{\text{mode}}_{j[t]} + \alpha^{\text{occasion}}_{k[t]}
\end{aligned}
$$ {#eq-mu}

where

$\beta_{\text{car},\, j[t]}$ denotes a coefficient of interaction between the N. of car and means of transportation (j=1, and 2; 1=walk, and 2=wheelchair); $\beta_{\text{person}}$ denotes a coefficient of the number of passengers; $\beta_{\text{motorcycle}}$ denotes a coefficient of the number of motorcycle; $\beta_{\text{lon}}$ denotes a coefficient of longitude; $\beta_{\text{lat}}$ denotes a coefficient of latitude; $\alpha^{\text{mode}}_{j[t]}$ denotes a fixed effect of mode $j$ ($j$=1,2; 1=walk, and 2=wheelchair) at time $t$; and $\alpha^{\text{occasion}}_{k[t]}$ denotes fixed effect of occasion $k$ ($k$=1,2,3; 1=morning, 2=afternoon, and 3=evening) at time $t$. To avoid $\log (0)$, we set .

Our analysis is based on Bayesian inference. Combining the state space model, the Bayesian inference enables us to run more flexible and complex model. For the analysis, we assume the following prior distributions for the target coefficients above (@eq-prior).

$$
\begin{aligned}
a &\sim \mathcal{N}(2, 1) \\
\beta_{\text{lon}}, \beta_{\text{lat}} &\sim \mathcal{N}(0, 1) \\
\beta_{\text{car}}, \beta_{\text{person}}, \beta_{\text{motorcycle}} &\sim \mathcal{N}(1, 0.5) \\
\alpha^{\text{mode}}_j &\sim \mathcal{N}(0, \sigma_{\text{mode}}), \quad \sigma_{\text{mode}} \sim \mathcal{N}^+(0, 1) \\
\alpha^{\text{occasion}}_k &\sim \mathcal{N}(0, \sigma_{\text{occasion}}), \quad \sigma_{\text{occasion}} \sim \mathcal{N}^+(0, 1) \\
\sigma &\sim \mathcal{N}^+(1, 0.5)
\end{aligned}
$$ {#eq-prior}


iteration

All the $\hat{r}$ were less than 1.05, suggesting that the model converged well.


We observe spatial coordinates \((y_{\text{lat}, n}, y_{\text{lon}, n})\) for \(n = 1, \dots, N\), with associated predictors:
- Mode: \(\text{mode}[n] \in \{1, \dots, J_{\text{mode}}\}\)
- Occasion: \(\text{occasion}[n] \in \{1, \dots, J_{\text{occasion}}\}\)


### Model to estimate moving location

For each observation $n$:

$$
\mathbf{y}_n = 
\begin{bmatrix}
\text{lat}_n \\\\ 
\text{lon}_n
\end{bmatrix}
\sim 
\mathcal{N}\left( 
\boldsymbol{\mu}_{\text{occasion}[n]} + \mathbf{b}_{\text{occasion}[n], \text{mode}[n]},
\ \boldsymbol{\Sigma} 
\right)
$$

where $\mathbf{y_{n}}$ denotes burabura; $\boldsymbol{\mu}_{j}$ is the fixed effect for occasion $j$; $\mathbf{b}_{j, m}$ denotes the random effect for mode $m$ nested within occasion $j$; and $\boldsymbol{\Sigma}$ is the residual covariance matrix of lat and lon.

We assume the prior distribution of fixed and random effect follow multivariate normal distributions given,

$$
\boldsymbol{\mu}_{j} \sim \mathcal{N}\left( 
\begin{bmatrix}
0 \\\\ 
0 
\end{bmatrix}, 
5^2 \mathbf{I}_2 
\right)
$$

and

$$
\mathbf{b}_{j, m} \sim \mathcal{N}\left( 
\begin{bmatrix}
0 \\\\ 
0 
\end{bmatrix}, 
2^2 \mathbf{I}_2 
\right)
$$

The residual covariance matrix is decomposed as:

$$
\boldsymbol{\Sigma} = \mathbf{L} \mathbf{L}^\top
$$

where $\mathbf{L}$ is $\operatorname{diag}(\boldsymbol{\sigma}) \times \mathbf{L}_{\text{corr}}$; and $\boldsymbol{\sigma}$ follows $\operatorname{Exponential}(1)$; and $\mathbf{L}_{\text{corr}}$ follows $\operatorname{LKJ}(2)$.

We implemented all the Bayesian models above in Stan (v2.33) via the cmdstanr interface (v0.6.0) and fitted using the Hamiltonian Monte Carlo (HMC) algorithm. For each model, we ran 4 independent chains with 1,000 warm-up iterations and 5,000 sampling iterations per chain (total: 20,000 post-warm-up samples). We monitored the potential scale reduction factor ($\hat{R}$) to assess chain convergence. All R-hat values were below 1.01, indicating satisfactory convergence. Effective sample sizes (n_eff) for all parameters exceeded 1,000, suggesting reliable posterior estimates. All model diagnostics ($\hat{R}$, effective sample size, energy plots, and trace plots) were inspected using the posterior and bayesplot packages in R (v4.3.1). No divergent transitions or energy Bayesian fraction of missing information (E-BFMI) issues were detected.

<!----- Results part from here ----->
# Results

## Estimating parapeters explaining moving speed

Table (@tbl-summary-main) indicates estimated posterior distribution of the key indicators listed in @eq-state, @eq-obs, @eq-mu, and @eq-prior.

We estimated the global intercept $\hat{\alpha}$ at 2.09 km/h (95% CI: 0.47–3.68), indicating a baseline level of moving speed across participants. 
<!-- Fast? slow? -->

Among covariates, the number of cars showed a statistically significant interaction with movement mode ($\hat{\beta}_{\text{car},\, j[t]}$). Specifically, for wheelchair users, an increase in the number of cars was associated with a slight but significant increase in moving speed ($\hat{β}$ = 0.018; 95% CI: 0.005–0.032), possibly reflecting avoidance behavior in response to traffic. Namely, when a wheelchair user happens to face on a car, to avoid accident, he/she tend to go faster.This effect was not observed among walkers. Effects of the number of persons and motorcycles were weak and statistically insignificant.





Group-level fixed effects for mode (walking vs. wheelchair) $\hat{\alpha}^{\text{mode}}_{j[t]}$ and occasion (morning, afternoon, evening) $\hat{\alpha}^{\text{occasion}}_{j[t]}$ were also non-significant, suggesting that temporal and spatial variations were better captured by the latent time-varying trend. The latent trend exhibited a strong positive autocorrelation ($\phi$ = 0.85; 95% CI: 0.82–0.87), indicating smooth transitions over time and persistent deviations in speed. Observation noise was moderate ($\hat{\sigma}_{y}$ = 0.316), while the variability of the latent trend was captured by σₓ = 0.472, highlighting the dynamic nature of movement speed beyond observable covariates.

```{r}
#| label: tbl-summary-main
#| tbl-cap: "Posterior summary of main parameters from AR(1) + Interaction Model"

summary_df <- read_csv("fit_moving_speed_spatial_ar1_interaction_summary.csv")

summary_main <- summary_df %>%
  filter(variable %in% c(
    "a", 
    "beta_car_by_mode[1]", "beta_car_by_mode[2]", 
    "beta_person", "beta_motorcycle",
    "alpha_mode[1]", "alpha_mode[2]",
    "alpha_occasion[1]", "alpha_occasion[2]", "alpha_occasion[3]",
    "sigma", "sigma_x", "phi1"
  )) %>%
  select(variable, mean, sd, `2.5%`, `97.5%`)

kbl(summary_main, 
    format = "latex", booktabs = TRUE, digits = 2) %>%
  kable_styling(latex_options = c("hold_position"))
```

## Where to move?




## Discussion


### Discussion

This study aimed to investigate temporal and contextual determinants of moving speed among pedestrians and wheelchair users in a tourism area, utilizing a Bayesian AR(1) + interaction model. Our results highlight the importance of accounting for both time-varying latent factors and interaction effects between movement mode and environmental covariates.

The estimated global intercept, $\hat{α}$ is approximately 2.09 km/h in speed, suggests a moderate baseline moving speed in the observed area, consistent with typical pedestrian speeds in urban settings. Notably, we identified a significant positive interaction between the number of cars and movement mode: for wheelchair users, an increase in nearby car traffic was associated with a slight but statistically significant increase in moving speed. This finding may reflect behavioral adaptations, such as efforts by wheelchair users to avoid moving obstacles or perceived hazards, leading to momentary acceleration.

<!-- loss of business opportunity -->

In contrast, the number of persons and motorcycles showed no meaningful association with speed after accounting for temporal trends and individual differences. Additionally, the mode-specific (walking vs. wheelchair) and occasion-specific (morning, afternoon, evening) effects were small and statistically non-significant, indicating that much of the variability in movement speed is better captured by the latent trend.

The latent trend, modeled as an AR(1) process, exhibited strong temporal persistence ($\phi$ is approximately 0.85$\pm$), suggesting that fluctuations in moving speed evolve smoothly over time and are influenced by unobserved contextual or behavioral factors. This autocorrelation underscores the necessity of dynamic modeling frameworks to capture underlying movement patterns beyond static predictors.

Overall, our findings emphasize the importance of considering temporal dependence and interaction effects in modeling pedestrian mobility, particularly when aiming to support urban planning, barrier-free accessibility, or crowd management in tourism areas. The presented framework may offer valuable insights for designing inclusive and responsive environments that accommodate diverse mobility needs.


車が現れると、車椅子は速度が上がる。車を避けようとするからと考えられる。
歩行者には車は影響しない。歩行者・車椅子双方に対して、他歩行者とバイクは影響しない。
なんで影響がないのか・なんで車椅子だけが車に影響されるか→歩くところが違うから。




路駐で商売あがったり→路駐しないように考えないと。
歩道を整備しよう




### Conclusion

<!-- hogehoge -->

Finally, we note several limitations of our study. First, although we introduced a dynamic trend component through the AR(1) process, the model does not account for potential spatial autocorrelation beyond longitude and latitude covariates. Future studies could extend the model to incorporate spatiotemporal dependencies explicitly. Second, the present analysis focused on aggregated numbers of cars, persons, and motorcycles without considering individual-level trajectories or interactions, which may overlook finer-scale behavioral responses. Third, while the interaction effect between mode and car traffic was statistically significant, the magnitude of the effect was relatively small, and further validation using external datasets is warranted. Finally, our analysis did not address the possible influence of environmental factors such as weather, road conditions, or pedestrian density beyond the immediate covariates included.
Conclusion

This study demonstrated that a dynamic regression model with an AR(1) latent trend and interaction terms provides an effective approach to capture temporal and contextual variability in pedestrian movement speed. The results suggest that wheelchair users may increase their moving speed in response to the presence of nearby cars, potentially reflecting adaptive behavior. The strong temporal autocorrelation of the latent trend underscores the relevance of unobserved, time-varying influences on mobility patterns. Our findings contribute to a deeper understanding of movement dynamics in tourism areas and highlight the potential of Bayesian time series modeling in supporting accessible and safe urban environments.


# References
